Ten Simple Rules for Teaching Programming

### 2017-07-07 paper.tex
Using paper.* for all names
Added a figure, plus various small fixes after a read-through.

Research from educational psychology suggests that teaching and learning are subject-specific
activities~\citep{mayer-subject}: learning programming has a different set of challenges and techniques
than learning physics or learning to read and write.  Computing is a younger discipline than
mathematics, physics, or biology, and while there have been correspondingly fewer studies of how best to teach it,
there is a growing body of evidence about what works and doesn't.
This paper presents ten simple rules that should be the foundation of any teaching, whether
formal or informal.

@@

\citet{guzdial-myths} refers to the belief that some people are born programmers and others aren't
as ``computing's most enduring and damaging myth.''
Our first and most important rule is that this is wrong:
competence at programming is not innate,
but is rather a learned skill that can be acquired and improved with practice.

@@

Beliefs such as this are known to have powerful effects on education outcomes
\cite{alvidrez-weinstein,brophy,jussim-eccles}.
If instructors believe that ``some kids get it and some kids don't'',
they will (consciously or unconsciously) invest less in those whom they put in the second category.
When combined with cultural stereotypes about who is and isn't a "natural programmer",
the downward spiral of under-achievement that results from differential attention
may be partly responsible for the gender imbalance in computing.

@@

No matter how good teachers are,
they can only say one thing at a time.
How then can they clear up learners' many different misconceptions
in a reasonable time?

The best method developed so far is called peer instruction.
Originally created by Eric Mazur at Harvard~\citep{mazur-peer-instruction},
it has been studied extensively in a wide variety of contexts,
including programming \cite{porter,cutts}
In simplified form,
peer instruction proceeds in three phases:

@@

\item The instructor then gives learners a multiple choice question
   that probes for misconceptions rather than simple factual recall.

@@

@@

Rather than using slides,
create programs in front of their learners~\cite{rubin-live-coding}.
This is more effective for multiple reasons:

@@

\item It enables instructors to be more responsive to ``what if?''
  questions. Where a slide deck is like a railway track, live coding
  allows instructors to go off road and follow their learners'
  interests.

\item It facilitates unintended knowledge transfer: students learn more
  than the instructor consciously intends to teach by watching \textit{how}
  instructors do things.

\item It slows the instructor down: if they has to type in the program
  as they go along, they can only go twice as fast as their
  learners, rather than ten-fold faster as they could with slides.

\item Learners get to how instructors diagnose and correct
  mistakes. Novices are going to spend most of their time doing
  this, but it's left out of most textbooks.

\item Watching instructors make mistakes shows learners that it's all
  right to make mistakes of their own~\citep{barker-live-coding}.
  Most people model the behavior of their teachers: if the instructor
  isn't embarrassed about making and talking about mistakes, learners
  will be more comfortable doing so too.

@@

\item Typing in boilerplate code that is needed by the lesson, but not
  directly relevant to it (such as library import statements), can
  distract learners from the intended thrust of a lesson.  As
  \citet{willingham-book} says, ``Memory is the residue of thought'';
  if the instructor spends their time typing boilerplate, that may be
  all that learners take away.

@@

When instructors are using live coding,
they usually run the program several times during its development to show what it does.
Research from peer instruction in physics education shows that
learners who observe a demonstration do not learn better than
those who did not see the demonstration~\citep{crouch-demo},
and in fact many learners misremember the outcome of demonstrations afterwards~\citep{miller-demo}.

The key to making demonstrations more effective is
to make learners predict the outcome of the demonstration before performing it.
Crucially, their prediction should be in some way recorded or public,
e.g. by a show of hands,
by holding up a cue card (A/B/C/D),
or by talking to their neighbour.
We speculate that the sting of possibly being publicly wrong leads learners to pay more attention
and to reflect on what they are learning;
regardless of whether this hypothesis is true,
instructors should be careful not to punish or criticise students who predicted wrongly,
but rather to use those incorrect predictions as a spur to further exploration.

@@

Pair programming is a software development practice
in which two programmers share one computer.
One person (called the driver) does the typing,
while the other (called the navigator) offers comments and suggestions.
The two switch roles several times per hour.
Pair programming is a good practice in real-life programming \cite{hannay},
and also a good way to teach \cite{mcdowell}.
Partners can not only help each other out during practical exercises,
but can also clarify each other's misconceptions when the solution is presented.

@@

A good idea when teaching is
to provide many similar examples for learners to practice on.
But since learners are novices,
they may not see the similarity between examples:
finding the highest rainfall from a list of numbers
finding the first surname alphabetically from a list of names
may seem like quite different problems to them.

\citet{morrison-subgoals,morrison-parsons} have shown that students perform better
when worked examples are broken down into steps (or subgoals)
which are given names (or labels).
Subgoal labels provide a structure which students can apply to future tasks that they attempt themselves,
and allow them to communicate with their peers and instructors more efficiently.

@@

A principle that applies across all areas of education is that transference only comes with mastery \cite{gick-holyoak}.
Courses should therefore stick to one language until learners have progressed far enough with it
to be able to distinguish the forest from the trees.
While an experience programmer can,
for example,
take what they know about loops and function calls in one language
and re-use that understanding in a language with a different syntax or semantics,
a newcomer does not yet know which elements of their knowledge are central
and which are accidental.
Attempting to force transference too early---i.e.,
requiring them to switch from Python to JavaScript in order to do a web programming course
early in their education---will confuse learners and erode their confidence.

@@

Guzdial et al.\ found that having learners manipulate images, audio, and video
in their early programming assignments
increased retention in two senses:
learners remembered more of the material when re-tested after a delay,
and were more likely to stay in computing programs \cite{guzdial-media}.
This is a particular instance of a larger observation:
learners find authentic tasks more engaging than abstracted examples.

@@

One caution is that context can inadvertently exclude some people while drawing others in.
For example,
many educators use computer games as a motivating example for programming classes,
but some learners may associate them with violence and racial or gender stereotypes.

@@

This principle sounds tautological, but it is easily forgotten.
Novices program differently than experts,
and need different approaches or tools.
If you ask a professional programmer to iterate over a list of integers and produce the average,
they can write the code within seconds,
using stored knowledge of the exact pattern required.
A novice will approach this problem totally differently:
they need to remember the syntax for the different parts,
they need to know how to iterate over a list,
how to use an accumulator variable,
and so on.

@@

\bibliographystyle{abbrvnat}
\bibliography{refs}

>>>>

Research from educational psychology suggests that teaching and learning are subject-specific
activities~\citep{mayer-subject}: learning programming has a different set of challenges and techniques
than learning physics or learning to read and write.  Computing is a younger discipline than
mathematics, physics, or biology, and while there have been correspondingly fewer studies of how best to teach it,
there is a growing body of evidence about what works and doesn't.
This paper presents ten simple rules that should be the foundation of any teaching of programming, whether
formal or informal.

@@

\citet{guzdial-myths} refers to the belief that some people are born programmers and others aren't
as ``computing's most enduring and damaging myth.''  This is often ``confirmed'' by looking at university grade distributions, which are commonly held to be bimodal: a low-scoring hump of those that will never get it, and a high-scoring hump of those that have the right stuff.
Our first and most important rule is that this is wrong:
competence at programming is not innate,
but is rather a learned skill that can be acquired and improved with practice.

@@

Beliefs such as this are known to have powerful effects on education outcomes
\citep{alvidrez-weinstein,brophy,jussim-eccles}.
If instructors believe that ``some kids get it and some kids don't'',
they will (consciously or unconsciously) invest less in those whom they put in the second category.
When combined with cultural stereotypes about who is and isn't a "natural programmer",
the downward spiral of under-achievement that results from differential attention
may be partly responsible for the gender imbalance in computing.

@@

One-on-one tutoring is perhaps the ideal form of teaching: all of a teacher's attention can be focused on one student, and they can completely customise their teaching for their single student, and tailor individual feedback and corrections based on two-way dialogue with that student.  In realistic settings however, one teacher must teach many students at once, whether that is a handful, tens or even hundreds in large lectures.  In larger settings, how can teachers possibly hope to clear up learners' many different misconceptions in a reasonable time?

The best method developed so far for larger-scale classrooms is called Peer Instruction.
Originally created by Eric Mazur at Harvard~\citep{mazur-peer-instruction},
it has been studied extensively in a wide variety of contexts,
including programming \citep{cutts,porter}.
In simplified form,
peer instruction proceeds in three phases:

@@

\item The instructor then gives learners a multiple choice question
   that probes for misconceptions rather than simple factual recall.  A programming example is given in figure \ref{fig:mcq} which relates to integer comparison and loops.

@@

\begin{figure}\sffamily
\begin{verbatim}
for (int i = 1; i < 10; i++)
{
     if (i < 3 || i > 8)
     {
          System.out.println("Yes");
     }
}
\end{verbatim}
How many times will the above code print out the word Yes?\\
a) 10\\
b) 9\\
c) 4\\
d) 3
\label{fig:mcq}
\caption{An example multiple choice question, probing learners' understanding of loops and integer comparisons.}
\end{figure}

@@

Rather than using slides,
instructors should create programs in front of their learners~\citep{rubin-live-coding}.
This is more effective for multiple reasons:

@@

\item It enables instructors to be more responsive to ``what if?''
  questions. Where a slide deck is like a highway, live coding
  allows instructors to go off road and follow their learners'
  interests.

\item It facilitates unintended knowledge transfer: students learn more
  than the instructor consciously intends to teach by watching \textit{how}
  instructors do things.  The extra knowledge may be high-level (e.g. whether a program is written top-down or bottom-up) or fairly low-level (e.g. using keyboard shortcuts).

\item It slows the instructor down: if the instructor has to type in the program
  as they go along, they can only go twice as fast as their
  learners, rather than ten-fold faster as they could with slides.

\item Learners get to see how instructors diagnose and correct
  mistakes. Novices are going to spend most of their time doing
  this, but it's left out of most textbooks.

\item Watching instructors make mistakes shows learners that it's alright to make mistakes of their own~\citep{barker-live-coding}.
  Most people model the behavior of their teachers: if the instructor
  isn't embarrassed about making and talking about mistakes, learners
  will be more comfortable doing so too.

@@

\item Instructors can spend too much time typing in boilerplate code that is needed by the lesson, but not
  directly relevant to it (such as library import statements), can
  distract learners from the intended thrust of a lesson.  As
  \citet{willingham-book} says, ``Memory is the residue of thought'';
  if the instructor spends their time typing boilerplate, that may be
  all that learners take away.  This can be avoided by starting with a partial skeleton that includes the boilerplate, or having it on hand to copy and paste at the appropriate juncture.

@@

When instructors are using live coding,
they usually run the program several times during its development to show what it does.
Research from peer instruction in physics education shows that
learners who observe a demonstration \textit{do not} learn better than
those who did not see the demonstration~\citep{crouch-demo},
and in fact many learners misremember the outcome of demonstrations afterwards~\citep{miller-demo}.  That is, demonstrations can be useless or actively harmful to learning.

The key to making demonstrations more effective is
to make learners predict the outcome of the demonstration before performing it.
Crucially, their prediction should be in some way recorded or public,
e.g. by a show of hands,
by holding up a cue card (A/B/C/D),
or by talking to their neighbour.
We speculate that the sting of being publicly wrong leads learners to pay more attention
and to reflect on what they are learning;
regardless of whether this hypothesis is true,
instructors should be careful not to punish or criticise students who predicted wrongly,
but rather to use those incorrect predictions as a spur to further exploration.

@@

Pair programming is a software development practice
in which two programmers share one computer.
One person (called the driver) does the typing,
while the other (called the navigator) offers comments and suggestions.
The two switch roles several times per hour.
Pair programming is a good practice in real-life programming \citep{hannay},
and also a good way to teach \citep{mcdowell}.
Partners can not only help each other out during practical exercises,
but can also clarify each other's misconceptions when the solution is presented.

@@

A good idea when teaching is
to provide many similar examples for learners to practice on.
But since learners are novices,
they may not see the similarity between examples:
finding the highest rainfall from a list of numbers and
finding the first surname alphabetically from a list of names
may seem like quite different problems to them, even though more advanced programmers would recognise them as isomorphic.

\citet{morrison-subgoals,morrison-parsons} have shown that students perform better
when worked examples are broken down into steps (or subgoals)
which are given names (or labels).
Subgoal labels provide a structure which students can apply to future tasks that they attempt themselves,
and allow them to communicate with their peers and instructors more efficiently.
TODO give an example in a figure.

@@

A principle that applies across all areas of education is that transference only comes with mastery \citep{gick-holyoak}.
Courses should therefore stick to one language until learners have progressed far enough with it
to be able to distinguish the forest from the trees.
While an experienced programmer can,
for example,
take what they know about loops and function calls in one language
and re-use that understanding in a language with a different syntax or semantics,
a newcomer does not yet know which elements of their knowledge are central
and which are accidental.
Attempting to force transference too early---e.g.,
requiring them to switch from Python to JavaScript in order to do a web programming course
early in their education---will confuse learners and erode their confidence.

@@

Guzdial et al.\ found that having learners manipulate images, audio, and video
in their early programming assignments
increased retention in two senses:
learners remembered more of the material when re-tested after a delay,
and were more likely to stay in computing programs \citep{guzdial-media}.
This is a particular instance of a larger observation:
learners find authentic tasks more engaging than abstracted examples.

@@

One caution is that context can inadvertently exclude some people while drawing others in.
For example,
many educators use computer games as a motivating example for programming classes,
but some learners may associate them with violence and racial or gender stereotypes, or simply find them unenjoyable.

This principle is tautological, but it is easily forgotten.
Novices program differently than experts,
and need different approaches or tools.
If you ask a professional programmer to iterate over a list of integers and produce the average,
they can write the code within seconds,
using stored knowledge of the exact pattern required.
A novice will approach this problem totally differently:
they need to remember the syntax for the different parts,
they need to know how to iterate over a list,
how to use an accumulator variable,
and so on.

@@

@@

\bibliographystyle{abbrvnat}
\bibliography{paper}
